---
title: "ST308 Project"
output:
  word_document: default
  html_notebook: default
  pdf_document: default
---

## About the data set

"https://www.kaggle.com/datasets/zynicide/wine-reviews?resource=download"

The data I've chosen to use is from 'Kaggle' and was scraped from 'WineEnthusiast'
during the week of June 15th, 2017. It contains data about: country, description, 
designation, points (quality), price, province, region_1, region_2, taster_name, 
taster_twitter_handle, title, variety, winery.


# --- Importing necessary libraries (and setting working directory) ---
```{r}
library(rstanarm)
library(rstan)
library(MASS)
library(glmnet)
library(ggplot2)
```

# --- Data cleaning ---

# Importing wine data
```{r}
wine <- read.csv("wine_data.csv", fileEncoding = "UTF-8")
colnames(wine)[c(1, 3, 5, 11, 12)] <- c("id", "desc", "rating", "twitter", 
                                        "review_title")
print("columns (variables):")
colnames(wine)
print("dimensions:")
dim(wine)
```

# Selecting chosen categories
```{r}
# assessing null or empty values 
na_or_empty_count <- sapply(wine, function(x) sum(is.na(x) | x == ""))
result <- data.frame(
  variables = names(na_or_empty_count),
  na_or_empty_count = as.vector(na_or_empty_count),
  stringsAsFactors = FALSE
)
result <- result[order(-result$na_or_empty_count), ]
print(result)

# selecting only necessary categories (which have sufficiently low levels of missing data)
wine <- subset(wine, select=c("country", "rating", "price", "province", "variety"))

# removing null or empty values
wine <- wine[apply(wine, 1, function(row) all(!is.na(row) & row != "")), ]

# creating a centered rating to speed up convergence
wine$rating <- as.numeric(as.character(wine$rating))
wine$rating_c <- wine$rating - mean(wine$rating)

# creating log_price to reduce the influence of higher priced outliers and make
# a more linear relationship with ratings
wine$price <- as.numeric(wine$price)
wine$log_price <- log2(wine$price) # base 2 chosen based on exploration of data

# creating lists of columns which are continuous, and categorical / discrete
cont_vars = c("price", "log_price", "rating", "rating_c")
cat_vars = c("country", "province", "variety")

# converting 'country', 'province', and 'region_1' to factors
for (var in cat_vars) {
  wine[[var]] <- factor(wine[[var]])
}

# removing provinces with less than 10 values
wine <- wine[table(wine$province)[wine$province] >= 10, ]
wine$province <- droplevels(wine$province)

# removing countries with less than 10 values
wine <- wine[table(wine$country)[wine$country] >= 10, ]
wine$country <- droplevels(wine$country)

# reviewing changes to data
print("chosen variables:")
colnames(wine)
print("new dimensions:")
dim(wine)
```

# --- Data Exploration (EDA) ---
```{r}
# data structure
str(wine)

# viewing summary stats for continuous variables
summary(wine[, cont_vars])
cat("\n")

# histogram of wine prices to show why log price is needed
hist(wine$price, freq = FALSE, col = "skyblue", border = "black",
     main = "Histogram of Wine Prices with Normal Curve",
     xlab = "Prices")
mean_price <- mean(wine$price)
sd_price <- sd(wine$price)
x_vals <- seq(min(wine$price), max(wine$price), length = 100)
y_vals <- dnorm(x_vals, mean = mean_price, sd = sd_price)
lines(x_vals, y_vals, col = "darkred", lwd = 2)
cat("The prices have mean average, ", mean_price, ", and standard deviation, ", 
    sd_price, ". ", "\n")

# histogram of log prices
hist(wine$log_price, freq = FALSE, col = "darkgreen", border = "black",
     main = "Histogram of Wine Log Prices with Normal Curve",
     xlab = "Log Prices")
mean_logprice <- mean(wine$log_price)
sd_logprice <- sd(wine$log_price)
x_vals <- seq(min(wine$log_price), max(wine$log_price), length = 100)
y_vals <- dnorm(x_vals, mean = mean_logprice, sd = sd_logprice)
lines(x_vals, y_vals, col = "navy", lwd = 2)
cat("The log prices seem to roughly follow a normal distribution around", 
    mean_logprice, "in the historgram, with a standard deviation of", 
    sd_logprice, ".")
```

```{r}
# viewing number of unique values of categorical variables
unique_counts <- sapply(cat_vars, function(var) {
  length(unique(wine[[var]]))
})
unique_counts_df <- data.frame(
  category = names(unique_counts),
  unique_count = as.integer(unique_counts),
  stringsAsFactors = FALSE
)
unique_counts_df <- unique_counts_df[order(-unique_counts_df$unique_count), ]
print(unique_counts_df)

# boxplot of log prices by country for relative comparison
boxplot(log_price ~ country, data = subset(wine),
        las = 2, col = "lightblue", main = "Wine (Log) Prices by Country")
```

```{r}
# removing variety as it has a very high count of unique values which will make interpretation more difficult
wine <- wine[, !names(wine) %in% "variety"]
cat_vars = c("country", "province")
dim(wine)
```

```{r}
# find countries with the highest level of data
country_counts <- table(wine$country)
top_countries <- sort(country_counts, decreasing = TRUE)[1:5]
print(top_countries)

# filter for US provinces and drop unused factor levels
us_data <- subset(wine, country == 'US')
us_data$province <- droplevels(us_data$province)
# create boxplot for only USA provinces
boxplot(log_price ~ province, data = us_data, col="red",
        las = 2, main = "Wine (Log) Prices by Region in the US", 
        xlab = "Province", ylab = "Wine Log Price")

# filter for French provinces and drop unused factor levels
france_data <- subset(wine, country == 'France')
france_data$province <- droplevels(france_data$province)
# create boxplot for only French provinces
boxplot(log_price ~ province, data = france_data, col="royalblue",
        las = 2, main = "Wine (Log) Prices by Region in France", 
        xlab = "Province", ylab = "Wine Log Prices")

# filter for Italian provinces and drop unused factor levels
italy_data <- subset(wine, country == 'Italy')
italy_data$province <- droplevels(italy_data$province)
# create boxplot for only Italian provinces
boxplot(log_price ~ province, data = italy_data, col="green",
        las = 2, main = "Wine (Log) Prices by Region in Italy", 
        xlab = "Province", ylab = "Wine Log Price")
```

# Lasso and Ridge regression modelling
```{r}
set.seed(42)

# creating design matrix (X) and response vector (y)
X <- model.matrix(log_price ~ rating_c + country + province, data = wine)[, -1]
y <- wine$log_price

# splitting data into training and test sets (80:20)
train_idx <- sample(seq_along(y), size = 0.8 * length(y))
X_train <- X[train_idx, ]
X_test  <- X[-train_idx, ]
y_train <- y[train_idx]
y_test  <- y[-train_idx]

# grid of lambda values
lambda_grid <- 10^seq(4, -2, length = 100)

# ridge
ridge_model <- cv.glmnet(X_train, y_train, alpha = 0, lambda = lambda_grid)
ridge_best_lambda <- ridge_model$lambda.min
cat("Ridge best lambda:", ridge_best_lambda, "\n")
ridge_pred <- predict(ridge_model, s = ridge_best_lambda, newx = X_test)
ridge_mse <- mean((ridge_pred - y_test)^2)
cat("Ridge test MSE:", ridge_mse, "\n\n")

# lasso
lasso_model <- cv.glmnet(X_train, y_train, alpha = 1, lambda = lambda_grid)
lasso_best_lambda <- lasso_model$lambda.min
cat("Lasso best lambda:", lasso_best_lambda, "\n")
lasso_pred <- predict(lasso_model, s = lasso_best_lambda, newx = X_test)
lasso_mse <- mean((lasso_pred - y_test)^2)
cat("Lasso test MSE:", lasso_mse, "\n")

# view number of selected coefficients for lasso
cat("\n", length(coef(lasso_model, s = lasso_best_lambda)[coef(lasso_model, s = lasso_best_lambda) != 0]), 
    "out of", length(coef(lasso_model, s = lasso_best_lambda)), "coefficients were selected (had non-zero values) by lasso.", "\n")
print(print(coef(lasso_model, s = lasso_best_lambda)[coef(lasso_model, s = lasso_best_lambda)[, 1] != 0, , drop = TRUE]))
```

# --- Creating models ---
```{r}
model_logprice <- stan_glmer(
  log_price ~ rating_c + (1 | country/province),
  data = wine,
  family = gaussian(),
  prior = normal(0, 1),
  prior_intercept = normal(mean(wine$log_price), 1),
  seed = 42,
  chains = 4,
  cores = 8,
  iter = 2000
)
```

```{r}
# saving the model
if (file.exists("model_logprice.rds") == FALSE) {
  saveRDS(model_logprice, file = "model_logprice.rds")
}
```

```{r}
# loading the model
if (file.exists("model_logprice.rds")) {
  model_logprice <- readRDS("model_logprice.rds")
  cat("Model loaded successfully.")
}
```

# --- Posterior Predictive Check ---
```{r}
pp_check(model_logprice)
```

# --- Convergence Diagnostics ---
```{r}
# check r-hat values (should be close to 1)
model_summary <- summary(model_logprice)
rhat_vals <- model_summary[, "Rhat"]
cat("\n", "summary of rhat values: ", "\n")
print(summary(rhat_vals))

# trace plots for 
plot(model_logprice, plotfun = "trace", pars = c("(Intercept)"))
plot(model_logprice, plotfun = "trace", pars = c("rating_c"))
plot(model_logprice, plotfun = "trace", pars = c("b[(Intercept) country:US]"))
plot(model_logprice, plotfun = "trace", pars = c("b[(Intercept) country:France]",
                                                 "b[(Intercept) country:Greece]",
                                                 "b[(Intercept) country:Italy]",
                                                 "b[(Intercept) country:Germany]"))
```

# --- Posterior Analysis ---
```{r}
options(max.print = 10000)

# summary of model, including credible intervals
print(summary(model_logprice, probs = c(0.05, 0.5, 0.95)), digits = 2)
```

# --- Examining random effects --- 
```{r}
# random effects
ranef_model <- ranef(model_logprice)
print(ranef_model)

# plotting variation in random intercepts for countries
country_effects <- ranef_model$country
country_df <- data.frame(country = rownames(country_effects), effect = 
                         country_effects[,1])
ggplot(country_df, aes(x = reorder(country, effect), y = effect)) +
       geom_point() +
       coord_flip() +
       labs(title = "Random Intercepts by Country", x = "Country", y = "Effect on 
            Log Price")

# plotting variation in random intercepts for province
province_effects <- ranef_model$province
print(province_effects)
us_province_effects <- subset(province_effects, )
province_df <- data.frame(province = rownames(province_effects), effect = 
                          province_effects[,1])
ggplot(province_df, aes(x = reorder(province, effect), y = effect)) +
       geom_point() +
       coord_flip() +
       labs(title = "Random Intercepts by Province", x = "Province", y = "Effect on 
            Log Price")
```

# --- Prior Summary ---
```{r}
prior_summary(model_logprice)
```
